Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
add freshness condition to Ctx-Cons. remove fig 2. adjust description of simultanous substitution. more description of typing rules.

In this section, we present the dependent type theory whose logical
consistency will be proven in Section~\ref{sec:logrelproof}. For conciseness,
we refer to this system as \lang.

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent universe types.%  and $[[x]]$
% represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
% $[[n]]$ are all metavariables representing natural numbers, we always use
% $[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
% variables.
% We use de Bruijn indices for our Coq development, but use nominal
% representation in the paper because it is more amenable to automation
% (Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
% conditions.
\yl{need to fix the text}
As a result, variables do not appear in binding forms, such as in
abstractions $[[ \ x . a ]]$ and dependent function types $[[Pi x : A . B]]$.  This latter
form binds a new variable of type $[[A]]$ that can appear in the codomain of
the type $[[B]]$.  We use the notation $[[A -> B]]$ when the output type
$[[B]]$ is not dependent on the input variable.
We include in \lang{} the intensional identity type
$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
$[[b]]$, and $[[c]]$ is the term whose type is to be casted.
Finally, \lang{} includes booleans, with standard syntax.
\scw{Example of
 what you can use identity types for? Or an example of a program that uses
J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
        [[idtm x]] & := & [[x]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho
        .: x -> a]]) } \\
        ([[rho .: x -> a]])([[x]]) & := & [[a]] \\
        ([[rho .: x -> a]])([[y]]) & := & [[rho y]] \text { if  } y
                                          \neq x\\ \\
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
        [[x { rho }  ]] & := &  [[rho x]] \\
        [[(Pi x : A . B) { rho }]] & := & [[Pi x : A { rho } . B { rho .: x -> x }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ x . a) { rho }]] & := & [[\ x . (a { rho .: x -> x })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

\yl{Need to fix the de bruijn text}
We adapt from \citet{autosubst2} notation for simultaneous renaming,
substitution, and other auxiliary definitions used in our paper, summarized in
Figure~\ref{fig:auxdef}. We need these definitions to specify the reduction
and typing relations for \lang{}. The metavariable $[[rho]]$ represents
substitutions, which are mappings from de Bruijn indices to terms.  One
example of a substitution is the function $\ottkw{id}_{tm}$,\scw{Can we just 
call this $\ottkw{id}$ now?} the identity
substitution that maps each index to the corresponding term variable.  % The
% $...$ substitution is more general and adds $[[n]]$ to the index
% before injecting it as a term.
The extension operation, $[[(rho .: x -> a)]]$,
allows us to update the substitution $[[rho]]$ to map the variable
$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.

The substitution operator, which takes the form $[[a {rho}]]$,
traverses the syntax of $[[a]]$ and replaces each variable $[[x]]$
with the term $[[rho x]]$. When traversing under binders (e.g. in the
$[[(\ x . a) { rho }]]$ case), we extend $[[rho]]$ to map the bound
variable $[[x]]$ to itself, protecting it from being replaced.
\scw{This substitution operation is pretty strange when it comes to the 
BVC. What variables are we allowed to use when we get to a bound variable $x$ in this function? I'd prefer if we instead had a side condition that it must be 
the case that $\rho(x) = x$ already, then we use BVC to pick such an $x$. But then we need to know that such an $x$ exists, and that isn't necessarily true.}
\scw{Maybe we should handwave here, and note that we aren't covering all of the 
details of the treatment of variable binding in the text. If readers want to understand that part, they should look at the Coq code, which uses de Bruijn indices anyways. Our goal is to convey the understanding of the Coq proof. }

@@

\yl{Explain why iterated substitution is not good enough}
\scw{resolved below, right?}

In our proofs, we find it
more convenient to formulate simultaneous substitution directly rather than
recovering it from single substitution.  In particular, this shows up in the
definition of semantic typing in Section~\ref{sec:logrelproof}, which relies
on simultaneous substitution.

@@

Our definition of equality is untyped: the judgement does not require the two
terms to type check and have the same type. The use of an untyped relation for
type conversion is reminiscent of Barendregt's Pure Type
Systems~\citet{barendregt1991introduction} and makes our formulation different
from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental
equality takes the form $\Gamma \vdash a \equiv b : A$, from which one can
usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
reduction. By working with an untyped judgement, we can establish its
properties independently from the type system and the logical relation, using
well-established syntactic approaches.

\citet{siles2012pure} shows the equivalence of Barendregt's Pure Type
System~\citep{barendregt1991introduction}, which employs untyped equality, and
its variant that uses typed judgmental equality. This assures us that we do
not lose generality working with a system with untyped
conversion.
\yl{removed the eta equivalence sentence because it might give the
  wrong message that we extended Siles' work with eta:
 Furthermore, discuss how this definition can be extended with
 $\eta$-equivalence of functions in Section~\ref{XXX} }
Finally, we compare this
definition with type-directed approaches to equality in
Section~\ref{sec:discuss}.

@@

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
 Void, Bool, True, False}
\[ \drule[width=5in]{T-If}\]
\[ \drule[width=3in]{T-Eq} \qquad \drule{T-Refl} \]
\[ \drule[width=4in]{T-J} \]
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

@@

\yl{Var, T-Pi, T-Abs are self-explanatory now?}
% In \rref{T-Var}, $G n$ is the partial function defined by the equations
% $ (G ++ A) 0 := A$ and $ (G ++ A) Suc n  := G n$.  The
% precondition $n < | G | $, where $ | G | $ represents the length of
% the context $G$, ensures that $G n$ has a defined output. The type
% $G n$ needs to be shifted by $Suc n$ since the types stored in the
% context $G$ are scoped differently. In the typing judgment
% $G |- a : A$, we want both $a$ and $A$ to have the same
% scope. However, the closer a type is to the head of $G$, the smaller de
% Bruijn it has since there is less variables in the context it can refer to
% (see \rref{Ctx-Cons}). Shifting $G n$ by $Suc n$ weakens the type so
% it is scoped consistently with the full context $G$ rather than a
% truncated version of $G$.

@@

The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
matching.  The result type of this expression $[[A {a / x}]]$ is composed of some
\emph{motive} $[[A]]$, a type where its single free variable has been replaced
with the condition of the if expression. When typing the true branch, this
substitution replaces the variable by $[[true]]$, and similarly for the false
branch. As a result, the type system communicates the information gained from
the % run-time \yl{commented out run-time per Jonathan's comment}
test to each of the branches of the expression.

A similar sort of dependent pattern matching occurs when eliminating identity
types. Such types are checked for well-formedness with \rref{T-Eq} and
introduced by \rref{T-Refl}. In \rref{T-J}, the elimination form, the subterm
$p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
is the body of the elimination form. In this rule, $[[B]]$ is the motive and has two free variables.  The substitution for these
variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
witnessing the information gained through dependent pattern matching.  In the
typing rule, the terms $[[a]]$ and $[[A]]$ need to be shifted in the context
for a similar reason; an extra type $[[A]]$ appears after the $[[G]]$ and
therefore the terms $[[a]]$ and $[[A]]$ in the identity type both need to be
shifted to refer to the same binders.

>>>>

In this section, we present the dependent type theory, \lang{}, whose logical
consistency will be proven in Section~\ref{sec:logrelproof}.

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent universe types, annotated by 
the universe level, a natural number $i$.%  and $[[x]]$
% represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
% $[[n]]$ are all metavariables representing natural numbers, we always use
% $[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
% variables.
% We use de Bruijn indices for our Coq development, but use nominal
% representation in the paper because it is more amenable to automation
% (Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
% conditions.
Abstractions $[[ \ x . a ]]$ and dependent function types $[[Pi x : A . B]]$
are binding forms for the variable $x$ in the body of the function and
codomain of the function type.  \footnote{In the exposition in this paper, we
  adopt the Barendregt Variable Convention and assume that binding forms are
  equal up to $\alpha$-conversion.  In some places we will be deliberately
  informal about the treatment of variables; our mechanized proofs use de
  Bruijn representation.}  We use the notation $[[A -> B]]$ when the output
type $[[B]]$ is not dependent on the input variable.  We include in \lang{}
the intensional identity type $[[a ~ b : A]]$ whose proofs can be eliminated
by the J-eliminator $[[J c a b p]]$, where $[[p]]$ is an equality proof
between $[[a]]$ and $[[b]]$, and $[[c]]$ is the term whose type is to be
casted.  Finally, \lang{} includes booleans, with standard syntax.
\scw{Example of what you can use identity types for? Or an example of a
  program that uses J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
%\begin{figure}[ht]
%  \begin{equation*}
%    \begin{split}
%      \begin{array}{lll}
%        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
%        [[idtm x]] & := & [[x]] \\ \\
%        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: x -> a]]) } \\
%        ([[rho .: x -> a]])([[x]]) & := & [[a]] \\
%        ([[rho .: x -> a]])([[y]]) & := & [[rho y]] \text { if  } y
%                                          \neq x\\ \\
%      \end{array}
%    \end{split}
%    \qquad \qquad
%    \begin{split}
%      \begin{array}{lll}
%        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
%        [[x { rho }  ]] & := &  [[rho x]] \\
%        [[(Pi x : A . B) { rho }]] & := & [[Pi x : A { rho } . B { rho .: x -> x }]] \\
%        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
%        [[(\ x . a) { rho }]] & := & [[\ x . (a { rho .: x -> x })]] \\
%        \ldots
%      \end{array}
%    \end{split}
%  \end{equation*}
%  \caption{Auxiliary Functions over Syntax}
%  \label{fig:auxdef}
% \end{figure}

@@

% We adapt from \citet{autosubst2} notation for simultaneous renaming,
% substitution, and other auxiliary definitions used in our paper, summarized in
% Figure~\ref{fig:auxdef}. 
Our reduction a typing relations are defined in terms of \emph{simultaneous
  substitutions}, $[[rho]]$, which are mappings from variables to terms.  We
use $[[idtm]]$ as the identity substitution.  The extension operation,
$[[(rho .: x -> a)]]$, updates the substitution $[[rho]]$ to map the variable
$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.

The substitution operator, which takes the form $[[a {rho}]]$, traverses the
syntax of $[[a]]$ and replaces each variable $[[x]]$ with the term
$[[rho x]]$. When traversing under binders (e.g. in the
$[[(\ x . a) { rho }]]$ case), it must be the case that $[[rho]]$ maps the
bound variable to itself and that the bound variable does not appear freely in
the application of the substitution to any other variable.
%
\scw{Maybe we should handwave more here, and note that we aren't covering all
  of the details of the treatment of variable binding in the text. If readers
  want to understand that part, they should look at the Coq code, which uses
  de Bruijn indices anyways. Our goal is to convey the understanding of the
  Coq proof. }

@@

When reasoning about logical relations, we find it more convenient to
formulate simultaneous substitution directly rather than recovering it from
single substitution.  In particular, this shows up in the definition of
semantic typing in Section~\ref{sec:logrelproof}, which relies on simultaneous
substitution.

@@

Our definition of equality is untyped: the judgement does not require the two
terms to type check and have the same type. The use of an untyped relation for
conversion is similar to Barendregt's Pure Type
Systems~\cite{barendregt1991introduction} and differs
from MLTT~\citep{Martin-Lof-1973}, where the judgmental
equality takes the form $\Gamma \vdash a \equiv b : A$.
% from which one can
% usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
% reduction. 
By working with an untyped judgement, we can establish its
properties independently from the type system and the logical relation, using
well-established syntactic approaches.
%
\citet{siles2012pure} show the equivalence of Barendregt's Pure Type
System, which employs untyped equality, and
its variant that uses typed judgmental equality. This assures us that we do
not lose generality working with a system with untyped
conversion.
\yl{removed the eta equivalence sentence because it might give the
  wrong message that we extended Siles' work with eta:
 Furthermore, discuss how this definition can be extended with
 $\eta$-equivalence of functions in Section~\ref{XXX} }
We compare this
definition with type-directed approaches to equality in
Section~\ref{sec:discuss}.

@@

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{}
\[ \drule{Ctx-Empty} \qquad \drule[width=3in]{Ctx-Cons} \]
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
 Void, Bool, True, False}
\[ \drule[width=5in]{T-If}\]
\[ \drule[width=3in]{T-Eq} \qquad \drule{T-Refl} \]
\[ \drule[width=4in]{T-J} \]
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

@@

\yl{Var, T-Pi, T-Abs are self-explanatory now?}  \scw{Need to actually mention
  these rules} The typing rules of \lang{} are standard rules for dependent
type theories.  The variable rule, \rref{T-Var}, uses the auxiliary relation
$[[x : A in G]]$, that holds when a variable declaration is found in the
typing context.  The typing of universes ensures that each one belongs to the
next higher level.  \Rref{T-Pi} ensures predicative quantification by
requiring that all parts of the type be typeable at the same universe
level. \Rref{T-Abs} ensures that all functions have well-formed dependent
types. In an application (\rref{T-App}) the argument is substituted for the
variable in the result type.

% In \rref{T-Var}, $G n$ is the partial function defined by the equations
% $ (G ++ A) 0 := A$ and $ (G ++ A) Suc n  := G n$.  The
% precondition $n < | G | $, where $ | G | $ represents the length of
% the context $G$, ensures that $G n$ has a defined output. The type
% $G n$ needs to be shifted by $Suc n$ since the types stored in the
% context $G$ are scoped differently. In the typing judgment
% $G |- a : A$, we want both $a$ and $A$ to have the same
% scope. However, the closer a type is to the head of $G$, the smaller de
% Bruijn it has since there is less variables in the context it can refer to
% (see \rref{Ctx-Cons}). Shifting $G n$ by $Suc n$ weakens the type so
% it is scoped consistently with the full context $G$ rather than a
% truncated version of $G$.

@@

The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
matching.  The result type of this expression, $[[A {a / x}]]$, is composed of some
\emph{motive} $[[A]]$, a type where its single free variable has been replaced
with the condition of the if expression. When typing the true branch, this
substitution replaces the variable by $[[true]]$, and similarly for the false
branch. As a result, the type system communicates the information gained from
the % run-time \yl{commented out run-time per Jonathan's comment}
test to each of the branches of the expression.

A similar sort of dependent pattern matching occurs when eliminating identity
types. Such types are checked for well-formedness with \rref{T-Eq} and
introduced by \rref{T-Refl}. In \rref{T-J}, the elimination form, the subterm
$p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
is the body of the elimination form. In this rule, $[[B]]$ is the motive and has two free variables.  The substitution for these
variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
witnessing the information gained through dependent pattern matching. 
% In the
% typing rule, the terms $[[a]]$ and $[[A]]$ need to be shifted in the context
% for a similar reason; an extra type $[[A]]$ appears after the $[[G]]$ and
% therefore the terms $[[a]]$ and $[[A]]$ in the identity type both need to be
% shifted to refer to the same binders.
